Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-06-14 doc/invt.tex
move old figures to fig1/

We now turn to a general multiple-bin updating scheme.
%
By definition, Eq. \eqref{eq:mbin_update},
a visit to bin $j$ in this case results in updates of the bias potential
not only at bin $j$, but also at a few neighboring bins $i$'s.
%
The optimization procedure is similar to the single-bin case.
%
The key step comes from the
projection of the bias potential to the $n$ eigenmodes
of the updating matrix, $\mathbf w$,
formed by the relative magnitudes, $w_{ij}$'s.

@@

The matrix, $\mathbf w$, is subject to three conditions.
%
First, we have a fixed-point condition\cite{bussi2006, dama2014}.
%
To sample the desired distribution,
$\mathbf p = (p_1, \dots, p_n)$,
the growth rate of the bias potential,
${\dot u}_i(t)$
in Eq. \eqref{eq:ut_diffeq_mbin},
should be a constant of $i$
if $h_j(t)$ were the same as $p_j$,
i.e.
$\sum_{j=1}^n w_{ij}$ should be a constant
to allow the bias potential to grow uniformly
in the asymptotic regime.
%
By a proper overall scaling of $\mathbf w$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
in our cases.
%
Below we will take advantage of the analogy
and borrow techniques used
in studying transition matrices\cite{vankampen}
(we will only use results that are unaffected by the allowance
of negative elements into $\mathbf w$).

@@

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
$$
{\tilde v}_k(t + 1) =
{\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
{\tilde \xi}_k(t)
,
$$
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde \xi}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \zeta}_k(t) \bigr]
  ,
  \label{eq:ut_diffeq_mbin}
\end{equation}
%
where we have removed the bin average,
and used the transformed version of Eq. \eqref{eq:sh_ave}
[but the subscript ``$_*$'' is dropped by Eq. \eqref{eq:ftstar}].
In the second step, we have used the linear approximation
as in Sec. \ref{sec:sbin_diffeq}.

@@

\subsection{\label{sec:lj}Lennard-Jones}

Our test system is a Lennard-Jones (LJ) fluid of $N=100$ particles,
and we wish to compute the PMF along the distance of two special particles, $r$.
%
The two special particles are mutually non-interacting
but they interact normally with other particles,\footnote{they
have placed on an axis parallel to a side of the cubic box}
and the intrinsic distribution, $p^*$, is
the cavity distribution function, $y(r)$\cite{hansen}.
%
We will assume reduced units in this calculation.
%
The reference bias potential was pre-calculated from a long simulation.

@@

The PMF was calculated from $r = 0$ to half of the box dimension
on a grid of size $\Delta r = 0.01$ in reduced units.
%
Two densities were tried, $\varrho^* = 0.1$ and $0.8$,
and the numbers of bins were $n = 500$ and $250$, respectively.
%
We tested the single-bin, Gaussian, and bandpass updating schemes.
%
Each simulation started with an ``equilibration'' phase
of $10^7$ steps under constant updating magnitude,
$a_0 = 10^{-4}$,\footnote{The equivalent $\ln f = n \, a_0$ for the WL algorithm
was $0.05$ and $0.025$, respectively.}
followed by a production phase
with updating magnitude given by
either the inverse-time or optimal schedule.
%
We repeated each simulation multiple times
and report the average.
%
To make sure the same schedule for multiple runs under the same condition,
the optimal schedule was pre-computed from a set of fixed values of
$\Gamma_k$ and $\epsilon_k$,
although they can also be fairly accurately estimated from the data
collected from the equilibration phase of each individual run.

As shown in Table 1,
both the inverse-time and optimal schedule reduced
the error significantly,
and the errors roughly matched the predicted values
calculated from the precomputed $\Gamma_k$'s and $\epsilon_k$'s.
%
The Gaussian scheme appeared to be superior
to the single-bin scheme in the $\varrho^* = 0.1$ case,
but not in the $\varrho^* = 0.8$ case.
%
In either case, there was an optimal width of Gaussian
(which was $\sigma = 0.37$ for $\varrho^* = 0.1$
or $0.07$ for $\varrho^* = 0.8$ in this case),
an overly large width could be counterproductive.

In Fig. \ref{fig:lj_xerr},
we show the error components against the eigenmode index
for the Gaussian updating scheme of width $\sigma = 0.2$.
%
While the Gaussian updating scheme effectively suppressed
the initial errors for short-range or large-$k$ modes,
it also slowed down the reduction of errors of these modes.
%
As a result, the final errors of the long-range or small-$k$ modes
were larger than those from the single-bin scheme.
%
For the Gaussian updating scheme,
the inverse-time schedule tended to reduce the updating magnitude
too quickly
leaving a hump on the error profile for the middle-range eigenmodes.
%
The optimal schedule improved over the inverse-time schedule
and produced a flatter profile for the final error.
%
It was also possible to roughly
recover the single-bin result by using the correction formula
based on the histogram, Eq. \eqref{eq:UHcorr}.
%
This correction usually benefited the small-$k$ modes
but not necessarily for the large-$k$ modes.

>>>>

We now turn to a general multiple-bin updating scheme.
%
By definition, Eq. \eqref{eq:mbin_update},
a visit to bin $j$ in this case results in updates of the bias potential
not only at bin $j$, but also at a few neighboring bins $i$'s.
%
Similar to Eq. \eqref{eq:ut_diffeq},
we can approximate the updating scheme
as a differential equation,
%
\begin{equation}
  \dot u_i(t)
  \approx
  \alpha(t) \,
  \sum_{j=1}^n w_{ij} \, \frac{ h_j(t) } { p_j }
  .
  \label{eq:ut_diffeq_mbin}
\end{equation}
%
The optimization procedure is similar to
the single-bin case.
%
The key step comes from the
projection of the bias potential to the $n$ eigenmodes
of the updating matrix, $\mathbf w$,
formed by the relative magnitudes, $w_{ij}$'s.

@@

The matrix, $\mathbf w$, is subject to three conditions.
%
First, we have a fixed-point condition\cite{bussi2006, dama2014}.
%
To sample the desired distribution,
$\mathbf p = (p_1, \dots, p_n)$,
the growth rate of the bias potential,
${\dot u}_i(t)$
in Eq. \eqref{eq:ut_diffeq_mbin},
should be a constant of $i$
if $h_j(t)$ were the same as $p_j$,
i.e.
$\sum_{j=1}^n w_{ij}$ should be a constant
to allow the bias potential to grow uniformly
in the asymptotic regime.
%
By a proper overall scaling of $\mathbf w$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative
in our cases.
%
Below we will take advantage of the analogy
and borrow techniques used
in studying transition matrices\cite{vankampen}
(we will only use properties that are unaffected
by the admission of negative elements into $\mathbf w$).

@@

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
$$
{\tilde v}_k(t + 1) =
{\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
{\tilde \xi}_k(t)
,
$$
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde \xi}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \zeta}_k(t) \bigr]
  ,
  \label{eq:vt_diffeq_mbin}
\end{equation}
%
where we have removed the bin average,
and used the transformed version of Eq. \eqref{eq:sh_ave}
[but the subscript ``$_*$'' is dropped by Eq. \eqref{eq:ftstar}].
In the second step, we have used the linear approximation
as in Sec. \ref{sec:sbin_diffeq}.
%
Equation \eqref{eq:vt_diffeq_mbin}
shows that the rate of error reduction
is now mode-dependent,
and the minimization of the final error
involves a compromise of the individual modes.

@@

%\subsection{\label{sec:lj}Lennard-Jones}

Our test system is a Lennard-Jones (LJ) fluid of $100$ particles,
and we wish to compute the PMF along the distance of two special particles, $r$.
%
The two special particles are mutually non-interacting
but they interact normally with other particles,\footnote{they
have placed on an axis parallel to a side of the cubic box}
and the intrinsic distribution, $p^*$, is
the cavity distribution function, $y(r)$\cite{hansen}.
%
We will assume reduced units in this calculation.
%
The reference bias potential was pre-calculated from a long simulation.

@@

The PMF was calculated from $r = 0$ to half of the box dimension
on a grid of size $\Delta r = 0.01$ in reduced units.
%
Two densities were tried, $\varrho^* = 0.1$ and $0.8$,
and the numbers of bins were $n = 500$ and $250$, respectively.
%
The reduced temperature was $T^* = 3$.
%
We tested the single-bin, Gaussian, and bandpass updating schemes.
%
Each simulation started with an ``equilibration'' phase
of $10^7$ steps under constant updating magnitude,
$a_0 = 10^{-4}$,\footnote{The equivalent $\ln f = n \, a_0$ for the WL algorithm
was $0.05$ and $0.025$, respectively.}
followed by a production phase of another $10^7$ steps
with updating magnitude given by
either the inverse-time or optimal schedule.
%
We repeated each simulation multiple times
and report the average.
%
To make sure the same schedule for multiple runs under the same condition,
the optimal schedule was pre-computed from a set of fixed values of
$\Gamma_k$ and $\epsilon_k$,
although they can be fairly accurately estimated from the data
collected from the equilibration phase of each individual run.

As shown in Table 1,
both the inverse-time and optimal schedule reduced
the error significantly,
and the errors roughly matched the predicted values
calculated from the precomputed $\Gamma_k$'s and $\epsilon_k$'s.
%
The Gaussian scheme appeared to be superior
to the single-bin scheme in the $\varrho^* = 0.1$ case,
but not in the $\varrho^* = 0.8$ case.
%
In the latter case,
the error was dominated by the $k=1$ eigenmode
corresponding to the longest range motion, as $\Gamma_1 \approx \Gamma$,
and the suppression of local noise
had little effect on reducing the total error.
%
In either case, there was an optimal width of Gaussian
an overly large width could be counterproductive.
%
The optimal width was $\sigma = 0.37$ for $\varrho^* = 0.1$
or $0.07$ for $\varrho^* = 0.8$ in this case,
although it depended on the simulation length.

To further investigate the behavior of the Gaussian updating scheme,
we show in Fig. \ref{fig:lj_xerr}
the error components, $\langle \tilde v_k^2 \rangle$,
against the eigenmode index
for the $\varrho^* = 0.1$ case
with width $\sigma = 0.2$.
%
While the Gaussian updating scheme effectively suppressed
the initial errors for short-range or large-$k$ modes,
it also slowed down the error reduction of these modes
[cf. Eq. \eqref{eq:vt_diffeq_mbin} and \eqref{eq:lambda_Gaussian}].
%
As a result, the final errors of the long-range or small-$k$ modes
were larger than those from the single-bin scheme.
%
For the Gaussian updating scheme,
the inverse-time schedule tended to reduce the updating magnitude
too quickly
leaving a hump on the error profile for the middle-range eigenmodes.
%
The optimal schedule improved over the inverse-time schedule
and produced a flatter profile for the final error.
%
It was also possible to roughly
recover the single-bin result by using the correction formula
based on the histogram, Eq. \eqref{eq:UHcorr}.
%
This correction usually benefited the small-$k$ modes
but not necessarily for the large-$k$ modes.
